IIT CS595: Topics \& Applications in Programming Languages\\ {\large Homework 1: Curry-Howard, Expressions continued}

### 2021-10-08 HW1/hw1.tex
hw1

\documentclass{article}
\usepackage[left=1in,right=1in,top=1in,bottom=1in]{geometry}
\usepackage{amsmath, amsthm, verbatim, enumerate, xcolor, mathpartir}

@@

@@

@@

@@

    \Rule{T-1}
         {\strut}
         {\etyped{\kwn{n}}{\kwmz}}
    \and
    \Rule{T-2}
         {n \neq 0}
         {\etyped{\kwn{n}}{\kwnz}}
    \and
    \Rule{T-3}
         {\etyped{e_1}{\kwmz}\\
           \etyped{e_2}{\kwmz}}
         {\etyped{e_1 + e_2}{\blank}}
    \and
    \Rule{T-4}
         {\etyped{e_1}{\kwmz}\\
           \etyped{e_2}{\kwmz}}
         {\etyped{e_1 - e_2}{\blank}}
    \and
    \Rule{T-5}
         {\etyped{e_1}{\kwnz}\\
           \etyped{e_2}{\kwnz}}
         {\etyped{e_1 * e_2}{\blank}}
    \and
    \Rule{T-6}
         {\etyped{e_1}{\kwmz}\\
           \etyped{e_2}{\kwmz}}
         {\etyped{e_1 * e_2}{\blank}}
    \and
    \Rule{T-7}
         {\etyped{e_1}{\kwmz}\\
           \etyped{e_2}{\blank}}
         {\etyped{e_1 / e_2}{\blank}}
  \end{mathpar}
}

@@

@@

@@

>>>>

\documentclass{article}
\usepackage[left=1in,right=1in,top=1in,bottom=1in]{geometry}
\usepackage{amsmath, amsthm, verbatim, enumerate, xcolor, mathpartir}
\usepackage{amsfonts} % for "\mathbb" macro

@@

\textbf{Answer:}
\kwforall {\alpha} {\kwforall {\beta} {\kwarr{(\kwsum{(\kwarr{\alpha}{\kwvoid})}{\beta})}({\kwarr{\alpha}{\beta}})}}

@@

\textbf{Answer:}
% this is probably not right...
\kwtfun{\alpha}{\kwtfun{\beta}{
    \\\hspace*{10mm} 
    \kwfun{b}{\beta}{
        \\\hspace*{15mm}
        \kwfun{l}{\kwsum{(\kwarr{\alpha}{\kwvoid})}{\beta}}{
            \\\hspace*{20mm} 
            \kwfun{a}{\alpha}{
                \\\hspace*{25mm} 
                \kwcase{l}
                    {x} {b}
                    {y} {y}}}}}}

When invoked with a value of type $\beta$, the above term will have the type specified in task 1.\\

@@

\textbf{Answer:}
    \kwtfun{\alpha}{
        \kwfun{f}{\kwarr {(\kwsum{\alpha}{(\kwarr{\alpha}{\kwvoid})})}{\kwvoid}}{
        \kwapp{f}{(\kwfun{a}{\kwsum{\alpha}{(\kwarr{\alpha}{\kwvoid})}}{
            \kwcase{a}{x}{\kwapp{f}{x}}{y}{\kwapp{f}{y}}})}}}

@@

    \Rule{T-1}
         {\strut}
         {\etyped{\kwn{n}}{\kwmz}}
    \and
    \Rule{T-2}
         {n \neq 0}
         {\etyped{\kwn{n}}{\kwnz}}
    \and
    \Rule{T-3}
         {\etyped{e_1}{\kwmz}\\
           \etyped{e_2}{\kwmz}}
         {\etyped{e_1 + e_2}{\kwmz}}
    \and
    \Rule{T-4}
         {\etyped{e_1}{\kwmz}\\
           \etyped{e_2}{\kwmz}}
         {\etyped{e_1 - e_2}{\kwmz}}
    \and
    \Rule{T-5}
         {\etyped{e_1}{\kwnz}\\
           \etyped{e_2}{\kwnz}}
         {\etyped{e_1 * e_2}{\kwnz}}
    \and
    \Rule{T-6}
         {\etyped{e_1}{\kwmz}\\
           \etyped{e_2}{\kwmz}}
         {\etyped{e_1 * e_2}{\kwmz}}
    \and
    \Rule{T-7}
         {\etyped{e_1}{\kwmz}\\
           \etyped{e_2}{\kwnz}}
         {\etyped{e_1 / e_2}{\kwmz}}
  \end{mathpar}
}

@@

\textbf{Answer:}
\begin{itemize}
    \item Rule (T-1): already of type \kwmz 
    \item Rule (T-2): $e = \kwn{n}$ thus apply (T-1)
    \item Rule (T-3): already of type \kwmz
    \item Rule (T-4): already of type \kwmz
    \item Rule (T-5): $e = e_1 * e_2$ \\
        by inversion $\etyped{e_1}{\kwnz}$ and $\etyped{e_2}{\kwnz}$\\
        by induction $\etyped{e_1}{\kwmz}$ and $\etyped{e_2}{\kwmz}$\\
        apply rule (T-6) giving $\etyped{e}{\kwmz}$
    \item Rule (T-6): already of type \kwmz
    \item Rule (T-7): already of type \kwmz
\end{itemize}

@@

\textbf{Progress:}
\begin{itemize}
    \item (T-1): $e = \kwn{n}$ and $\tau = \kwmz$ \\
        according to (V-1) $e \val$.
    \item (T-2): $e = \kwn{n}$ and $\tau = \kwnz$ \\
        according to (V-1) $e \val$.
    \item (T-7): $e = e_1 / e_2$ and $\tau = \kwmz$ \\
        by inversion $\etyped{e_2}{\kwnz}$\\
        by inversion on (T-2) $e_2 \neq 0$\\
        by induction $e_1$ and $e_2$ either step or are values\\ 
        if they are values we can use the canonical forms lemma to show that e steps via (S-6)\\
        if they step then the expression steps, proving the proposition.
\end{itemize}

\textbf{Preservation:}
\begin{itemize}
    \item (S-3): $e = e_1 + e_2$ and $\tau = \kwmz$\\
        by inversion on (T-3) $\etyped{e_1}{\kwmz}$ and $\etyped{e_2}{\kwmz}$
        (note that theorem 1 says this includes \kwnz)\\
        thus by induction on rule (T-3) $\etyped{e'}{\kwmz}$
    \item (S-4): $e = e_1 - e_2$ and $\tau = \kwmz$\\
        by inversion on (T-3) $\etyped{e_1}{\kwmz}$ and $\etyped{e_2}{\kwmz}$
        (note that theorem 1 says this includes \kwnz)\\
        thus by induction on rule (T-3) $\etyped{e'}{\kwmz}$
    \item (S-5): $e = e_1 * e_2$
        there are two cases to consider
        \begin{enumerate}
            \item when $\etyped{e_1}{\kwnz}$ and $\etyped{e_2}{\kwnz}$, $\tau = \kwnz$\\
                by inversion on (T-5) $\etyped{e_1}{\kwnz}$ and $\etyped{e_2}{\kwnz}$\\
                thus by induction $\etyped{e}{\kwnz}$
            \item when $\etyped{e_1}{\kwmz}$ or $\etyped{e_2}{\kwmz}$, $\tau = \kwmz$\\
                by inversion on (T-6) $\etyped{e_1}{\kwmz}$ and $\etyped{e_2}{\kwmz}$ 
                (note that theorem 1 says this includes \kwnz)\\
                thus by induction $\etyped{e'}{\kwmz}$
        \end{enumerate}
    \item (S-6): $e = e_1 / e_2$ and $\tau = \kwmz$\\
        by inversion on (T-7) $\etyped{e_1}{\kwmz}$ and $\etyped{e_2}{\kwnz}$\\
        thus by induction $\etyped{e'}{\kwmz}$
\end{itemize}

@@

\textbf{Example Expression:} $555 / (1 + 4)$ \\
We could extend the type system, allowing in addition to tracking zero-ness we also track the signedness of numbers. So we could have some new addition rules like the following shorthand for the type rules where u* means unsigned ($\ge 0$) and s* means signed ($\mathbb{Z}$)

\begin{enumerate}
    \item umz + umz : umz
    \item umz + unz : unz
    \item unz + umz : unz
    \item unz + unz : unz\\

    \item smz + smz : smz
    \item smz + snz : smz
    \item snz + smz : smz
    \item snz + snz : smz\\

    \item depending on implementation, it might also make sense to add negative signed types
\end{enumerate}

So in the example we could have used rule 4 to verify that the denominator would not be zero. This would still not work with all valid expressions but at least this could potentially be done at compile time.

Also partial evaluation could potentially help.
